Simulating International Energy Security

### 2014-01-19 Writeups/SBP Paper/Paper.tex
Adding SBP conference paper

\documentclass{article}
	\usepackage{natbib}
	\usepackage{mathtools}
	\usepackage{amsmath}
	\usepackage{algorithm2e}
	\usepackage{graphicx}
	\usepackage{subcaption}
	\usepackage{setspace}

@@

\onehalfspacing

@@

Energy security is defined by the International Energy Agency as ``the uninterrupted availability of energy sources at an affordable price''\citep{iea_2013}. As the world's energy demand continues to expand in the face of growing scarcity, energy security is becoming increasingly critical not just to developed countries but to developing ones as well \citep{yergin_2006}. Thus, assessing the energy security of different countries is important in order to understand the overall state of the international system. Despite this, there have been few attempts to create formal models of energy security.

Much of the analysis to date has taken a portfolio-based approach. \citet{wu_2009} uses portfolio diversity to analyze China's mix of crude oil and processed petroleum products. The paper notes that ``Oil import-specific risk derives from events, such as political events, natural disaster and so on, that are not related to general movements in the international market'', but does not analyze the risks of such events directly. It does, however, address the relatively less significant risk of piracy along several shipping routes. \citet{stirling_2010} and \citet{skea_2010} take a broader, more theoretical approach, and develops a multicriteria methodology for evaluating the diversity of an entire energy portfolio as it contributes to energy security. However, changing the composition of an entire energy portfolio takes a great deal of time -- it is a long-term process, measured in years or decades. Similarly, \citet{jacobson_2009} evaluates a set of possible future courses of action to jointly address climate, environmental and energy-security concerns -- again, courses of actions only implementable over the long term.

Short-term energy security, in contrast, appears to have been little-addressed in the literature. However, an important short-term analysis, MOSES (Model of Short-term Energy Security), was created within the International Energy Agency (IEA), the intergovernmental organization of the world's major oil importers established in part to protect their energy security. As the name of the model indicates, MOSES focuses on short-term risks, which it defines as a timescale of days and weeks. Furthermore, MOSES assigns risk scores based on a broad but ultimately static set of indicators. In particular, the indicator used to measure political stability is the ``weighted average of political stability of suppliers''. However, the greater the diversity of suppliers (another input into the model), the greater the potential range of political stabilities. Averaging ratings of political risk eliminates the complexity that characterizes international political systems \citep{cederman_1997,geller_2011}, and ignores the possibility of crisis and conflict contagion \citep{black_2013}, by which conflict in one place can trigger other conflicts nearby.

In fact, as \citet{hamilton_2003} describes, oil supply shocks have been driven caused by conflict involving oil producers, both domestic and international. \citet{hamilton_2003} argues that supply shock-driven changes to oil prices have a nonlinear effect on GDP, while \citet{kilian_2008} argues that the macroeconomic effects of the oil shocks themselves are small. Nevertheless, the price of oil is only an indirect indicator of energy security; if a substantial volume of oil cannot be extracted or exported, and hence is unavailable for consumption, it poses an energy security risk regardless of what the price of oil is doing.

@@

The network of oil imports and exports comes from the United Nations-maintained COMTRADE database \citep{un_2013}. I extract all imports of crude petroleum (product code 2709) and other unprocessed petroleum (product code 2710) for 2012, the most recent year for which complete data is available. Note that import data appears to be more reliable than export data: many major oil producers (particularly Saudi Arabia and Venezuela) only report export volume by region (e.g. Europe, North America); however, the majority of their trade partners report their imports from them.

For the majority of countries, I assume that supply and demand are equal to total export and imports, respectively. For the top ten oil producers, I use US Energy Information Administration (EIA) estimates \citep{eia_2013}, country estimates \citep{canada_2009,eia_domestic}, expert reports \citep{mohamedi_2010}, and media sources \citep{rasmi_2013} for those countries' consumption of their own domestic production.

Spare capacity is particularly difficult to estimate, and varies over time based on the price of oil and other factors \citep{mearns_2012}. I use media sources such as \citet{daya_2012} and subject-matter expert (SME) analysis such as \citet{mearns_2012} in order to estimate the spare capacity of other countries, when possible. I also assign a default spare capacity of 10\% based on SME input.

Political instability estimates are taken from the Economist Intelligence Unit's Political Index ratings \citep{eiu_2013}. These are in turn estimated based on the methodology developed by the Political Instability Task Force \citep{goldstone_2005}, which could predict the onset of internal instability in a two-year period with 80\% accuracy. Thus, these scores are normalized such that a rating of 10 (the maximum) is associated with an approximately 80\% probability of instability within a 24-month timeframe, translating into a 6.5\% monthly probability of crisis onset.

Crisis duration is drawn independently from onset. \citet{cioffi_2004} and others have argued that internal and external conflict durations follow a power law distribution, which I use here. I calibrate the power law coefficient from two datasets: the UCDP/PRIO Armed Conflict Dataset \citep{lotta_2013} for military conflicts, and the Social Conflict in Africa Database \citep{hendrix_2013}, which includes both armed and lower-level social conflicts. I subset the PRIO dataset for conflicts in Africa and merge it with SCAD, eliminating duplicates and events below the severity of a general strike, in order to obtain as complete as possible a set of crises and conflicts that may be associated with an oil production  shock. I find the duration of each event in months, and fit a power law to the resulting distribution, resulting in an estimated coefficient of $\mathbf{-1.37}$.

@@

The model was written in Java, using the MASON package \citep{luke_2005}, with additional analysis done in Python.

@@

\paragraph{Oil Submodel}
\begin{itemize}
	\item \textbf{Domestic Share} of the country's oil demand satisfied by domestic production. For most countries, this defaults to 0; for the top ten oil producers, it is set based on US Energy Information Administration data \citep{eia_2013}.
	\item \textbf{Total Imports} is the sum of the current volume of all oil imports. 
	\item \textbf{Total Demand} is the sum of current imported and domestic oil consumed. Since the EIA reports the share of total consumption that is domestically-sourced, demand is calculated as: $$\text{Total Demand}_i = \sum_j e_{j,i} \cdot (1 - \text{Domestic Share})^{-1}$$
	With $Total Imports$ set at the initial baseline. If current imports increase (as explained below) then supply will exceed demand. 
	\item \textbf{Total Exports} is the sum of current exported oil to all other countries.
\end{itemize}

\paragraph{Crisis Submodel}
\begin{itemize}
	\item \textbf{Instability} is a measure of the country's political instability on a 0-10 scale, and determines the probability of a crisis or conflict -- a production shock. It is obtained from \citet{eiu_2013}. 
	\item \textbf{In Crisis} is simply a boolean variable determining whether or not a country is currently experiencing a crisis or conflict.
	\item \textbf{Crisis Length} is the number of months remaining in a crisis, if one is currently in process. When a country enters crisis, this number is drawn from a power law distribution. Specifically, in order to accommodate the limitations of the MASON random-number generator, I use a uniform approximation \citep{weisstein_2013}, defined as:
$$
\left((x_{1}^{\alpha} - x_{0}^{\alpha})y + x_{0}^{\alpha}\right)^{\frac{1}{\alpha}} \sim \text{Power Law}
$$
where 
$$r.v. y \sim U[0,1]$$
$$ x_0  \equiv \text{minimum}$$
$$ x_1  \equiv \text{maximum}$$
$$ \alpha \equiv \text{Coefficient}$$

@@

\paragraph{Assistance Submodel}
\begin{itemize}
	\item \textbf{Spare Capacity} is the country's capacity to rapidly increase oil for export above its initial baseline. Defaults to 0.1 (that is, countries can rapidly increase output by 10\%) based on estimation from \citet{mearns_2012,eia_opec} and  consultations with subject-matter experts. 
	\item \textbf{Increasing Production} is a boolean of whether or not the country is currently in a state of temporarily-increased production.
	\item \textbf{Increasing Production For} stores the country being assisted. I assume that countries target their assistance at one other country at any given time.

@@

\begin{figure}[h!]
	\centering
	\begin{subfigure}{\textwidth}
		\includegraphics[width=\textwidth]{../Graphics/ContagionDistribution}
		\caption{Effect of Contagion}
	\end{subfigure}

	\begin{subfigure}{\textwidth}
		\includegraphics[width=\textwidth]{../Graphics/AssistanceDistribution}
		\caption{Effect of Assistance}
	\end{subfigure}

	\caption{Distribution of Overall Ratios with model parameters turned on and off}
\end{figure}

@@

\begin{figure}[h!]
	\centering
	\begin{subfigure}{\textwidth}
		\includegraphics[width=\textwidth]{../Graphics/SensitivityAnalysisOverall}
		\caption{Overall Ratio}
	\end{subfigure}

	\begin{subfigure}{\textwidth}
		\includegraphics[width=\textwidth]{../Graphics/SensitivityAnalysisSupply}
		\caption{Supply Ratio}
	\end{subfigure}

	\caption{Sensitivity analysis of ratio means and variances}
\end{figure}

@@

\singlespacing
\bibliographystyle{chicago}
\bibliography{energysecurity}

>>>>

\documentclass{llncs}
	\usepackage{mathtools}
	\usepackage{amsmath}
	\usepackage{algorithm2e}
	\usepackage{graphicx}
	\usepackage{setspace}

@@

@@

Energy security is defined by the International Energy Agency as ``the uninterrupted availability of energy sources at an affordable price''\cite{iea_2013}. As the world's energy demand continues to expand in the face of growing scarcity, energy security is becoming increasingly critical not just to developed countries but to developing ones as well \cite{yergin_2006}. Thus, assessing the energy security of different countries is important in order to understand the overall state of the international system. Despite this, there have been few attempts to create formal models of energy security.

Much of the analysis to date has taken a portfolio-based approach. \cite{wu_2009} uses portfolio diversity to analyze China's mix of crude oil and processed petroleum products. The paper notes that ``Oil import-specific risk derives from events, such as political events, natural disaster and so on, that are not related to general movements in the international market'', but does not analyze the risks of such events directly. It does, however, address the relatively less significant risk of piracy along several shipping routes. \cite{stirling_2010} and \cite{skea_2010} take a broader, more theoretical approach, and develops a multicriteria methodology for evaluating the diversity of an entire energy portfolio as it contributes to energy security. However, changing the composition of an entire energy portfolio takes a great deal of time -- it is a long-term process, measured in years or decades. Similarly, \cite{jacobson_2009} evaluates a set of possible future courses of action to jointly address climate, environmental and energy-security concerns -- again, courses of actions only implementable over the long term.

Short-term energy security, in contrast, appears to have been little-addressed in the literature. However, an important short-term analysis, MOSES (Model of Short-term Energy Security), was created within the International Energy Agency (IEA), the intergovernmental organization of the world's major oil importers established in part to protect their energy security. As the name of the model indicates, MOSES focuses on short-term risks, which it defines as a timescale of days and weeks. Furthermore, MOSES assigns risk scores based on a broad but ultimately static set of indicators. In particular, the indicator used to measure political stability is the ``weighted average of political stability of suppliers''. However, the greater the diversity of suppliers (another input into the model), the greater the potential range of political stabilities. Averaging ratings of political risk eliminates the complexity that characterizes international political systems \cite{cederman_1997,geller_2011}, and ignores the possibility of crisis and conflict contagion \cite{black_2013}, by which conflict in one place can trigger other conflicts nearby.

In fact, as \cite{hamilton_2003} describes, oil supply shocks have been driven caused by conflict involving oil producers, both domestic and international. \cite{hamilton_2003} argues that supply shock-driven changes to oil prices have a nonlinear effect on GDP, while \cite{kilian_2008} argues that the macroeconomic effects of the oil shocks themselves are small. Nevertheless, the price of oil is only an indirect indicator of energy security; if a substantial volume of oil cannot be extracted or exported, and hence is unavailable for consumption, it poses an energy security risk regardless of what the price of oil is doing.

@@

The network of oil imports and exports comes from the United Nations-maintained COMTRADE database \cite{un_2013}. I extract all imports of crude petroleum (product code 2709) and other unprocessed petroleum (product code 2710) for 2012, the most recent year for which complete data is available. Note that import data appears to be more reliable than export data: many major oil producers (particularly Saudi Arabia and Venezuela) only report export volume by region (e.g. Europe, North America); however, the majority of their trade partners report their imports from them.

For the majority of countries, I assume that supply and demand are equal to total export and imports, respectively. For the top ten oil producers, I use US Energy Information Administration (EIA) estimates \cite{eia_2013}, country estimates \cite{canada_2009,eia_domestic}, expert reports \cite{mohamedi_2010}, and media sources \cite{rasmi_2013} for those countries' consumption of their own domestic production.

Spare capacity is particularly difficult to estimate, and varies over time based on the price of oil and other factors \cite{mearns_2012}. I use media sources such as \cite{daya_2012} and subject-matter expert (SME) analysis such as \cite{mearns_2012} in order to estimate the spare capacity of other countries, when possible. I also assign a default spare capacity of 10\% based on SME input.

Political instability estimates are taken from the Economist Intelligence Unit's Political Index ratings \cite{eiu_2013}. These are in turn estimated based on the methodology developed by the Political Instability Task Force \cite{goldstone_2005}, which could predict the onset of internal instability in a two-year period with 80\% accuracy. Thus, these scores are normalized such that a rating of 10 (the maximum) is associated with an approximately 80\% probability of instability within a 24-month timeframe, translating into a 6.5\% monthly probability of crisis onset.

Crisis duration is drawn independently from onset. \cite{cioffi_2004} and others have argued that internal and external conflict durations follow a power law distribution, which I use here. I calibrate the power law coefficient from two datasets: the UCDP/PRIO Armed Conflict Dataset \cite{lotta_2013} for military conflicts, and the Social Conflict in Africa Database \cite{hendrix_2013}, which includes both armed and lower-level social conflicts. I subset the PRIO dataset for conflicts in Africa and merge it with SCAD, eliminating duplicates and events below the severity of a general strike, in order to obtain as complete as possible a set of crises and conflicts that may be associated with an oil production  shock. I find the duration of each event in months, and fit a power law to the resulting distribution, resulting in an estimated coefficient of $\mathbf{-1.37}$.

@@

The model was written in Java, using the MASON package \cite{luke_2005}, with additional analysis done in Python.

@@

\paragraph{Oil Submodel}
\begin{itemize}
	\item \textbf{Domestic Share} of the country's oil demand satisfied by domestic production. For most countries, this defaults to 0; for the top ten oil producers, it is set based on US Energy Information Administration data \cite{eia_2013}.
	\item \textbf{Total Imports} is the sum of the current volume of all oil imports. 
	\item \textbf{Total Demand} is the sum of current imported and domestic oil consumed. Since the EIA reports the share of total consumption that is domestically-sourced, demand is calculated as: $$\text{Total Demand}_i = \sum_j e_{j,i} \cdot (1 - \text{Domestic Share})^{-1}$$
	With $Total Imports$ set at the initial baseline. If current imports increase (as explained below) then supply will exceed demand. 
	\item \textbf{Total Exports} is the sum of current exported oil to all other countries.
\end{itemize}

\paragraph{Crisis Submodel}
\begin{itemize}
	\item \textbf{Instability} is a measure of the country's political instability on a 0-10 scale, and determines the probability of a crisis or conflict -- a production shock. It is obtained from \cite{eiu_2013}. 
	\item \textbf{In Crisis} is simply a boolean variable determining whether or not a country is currently experiencing a crisis or conflict.
	\item \textbf{Crisis Length} is the number of months remaining in a crisis, if one is currently in process. When a country enters crisis, this number is drawn from a power law distribution. Specifically, in order to accommodate the limitations of the MASON random-number generator, I use a uniform approximation \cite{weisstein_2013}, defined as:
$$
\left((x_{1}^{\alpha} - x_{0}^{\alpha})y + x_{0}^{\alpha}\right)^{\frac{1}{\alpha}} \sim \text{Power Law}
$$
where $r.v. y \sim U[0,1]$, $ x_0  \equiv \text{minimum}$, $ x_1  \equiv \text{maximum}$, and $ \alpha \equiv \text{Coefficient}$

@@

\paragraph{Assistance Submodel}
\begin{itemize}
	\item \textbf{Spare Capacity} is the country's capacity to rapidly increase oil for export above its initial baseline. Defaults to 0.1 (that is, countries can rapidly increase output by 10\%) based on estimation from \cite{mearns_2012,eia_opec} and  consultations with subject-matter experts. 
	\item \textbf{Increasing Production} is a boolean of whether or not the country is currently in a state of temporarily-increased production.
	\item \textbf{Increasing Production For} stores the country being assisted. I assume that countries target their assistance at one other country at any given time.

@@

@@

@@

\bibliographystyle{splncs}
\bibliography{energysecurity}
